02_models_1

### 2025-04-07 02_models_1.Rmd
first testing of the whole flow with all species with temp GAM

# function for the temperature only GAMs
source(here("function_GAM.R"))

# function for temp and depth models 
source(here("function_temp_depth.R"))
```

>>>>

# function for the temperature only GAMs with all run as expected
source(here("predictions_function.R"))
```

```{r}
# prepare folders

main_folder <- "gams_testing_1"
if (!dir.exists(main_folder)){
  dir.create(main_folder)}

subfolders <- c("gam_check", "predictions", "derivatives", "resp_div")

for (sub in subfolders){
  sub_path <- file.path(main_folder, sub)
  if(!dir.exists(sub_path)){
    dir.create(sub_path)
  }
}

# read in the dataframes
df_1 <- readRDS("data_frame_models/df_binomial_gam")
df_2 <- readRDS("data_frame_models/df_abundance_gam")
df_3 <- readRDS("data_frame_models/df_binomial_re")
df_4 <- readRDS("data_frame_models/df_abundance_re")

predictions(df_1)
predictions(df_2)
predictions(df_3)
predictions(df_4)

total_models <- bind_rows(df_1, df_2, df_3, df_4)

stats_predictions(total_models)

df_predictions_all <- list.files(path = "gams_testing_1/predictions", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

# save total predictions as RDS
saveRDS(df_predictions_all, "gams_testing_1/df_pred_all.rds")

```

####Fig. 2
Overview of all model predictions

```{r, fig.show='hide'}
# Load total predictions
model_predictions <- readRDS("gams_testing_1/df_pred_all.rds")

# check the species list
model_predictions$species <- as.factor(model_predictions$species)
levels(model_predictions$species)

# read in the category table and rename categories
species_across_lakes <-  read_xlsx("species_category_across_lakes.xlsx") |> 
  select(-notes) |> 
    mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
  mutate(category2 = factor("all"))

species_across_lakes$category <- as.factor(species_across_lakes$category)
levels(species_across_lakes$category)

#merge predictions with categories
predictions_categories <- merge(model_predictions, species_across_lakes)

# EVTL NEW PLOT 2
# change category names for facet_wrap
category_names <- c(
  `endemic` = "endemic",
  `endemic_and_translocated` = "translocated endemic",
  `non_endemic_native` = "non-endemic native",
  `non_endemic_native_and_translocated` = "translocated non-endemic native",
  `non_native` = "non-native"
)

# colors for the plot
mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")

# Rescale the predictions to 0-1 
predictions_categories_rescaled <- predictions_categories |>
  group_by(species) |>
  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))

plot_pred <- predictions_categories_rescaled |>
  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
  geom_line() +
  # how can I get the right color?
   # geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), alpha = 0.3) +
  theme_bw(base_size = 16) +
  ylab("Abundance") +
  xlab("Temperature") +
  scale_color_manual(values = mycolors1, guide = NULL) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "bottom",
        legend.title = element_blank())

plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))

plot_category_predictions

# trying to group by category and color
figure_test_2 <- predictions_categories |>
  ggplot(aes(temp, fit)) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(category)), alpha = 0.3) +
  geom_line(aes(color = factor(category))) +
  theme_bw() +
  facet_wrap(~species + category, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
  scale_color_manual(values = mycolors1, aesthetics = c("color", "fill")) +
  xlab("Temperature") +
  ylab("Abundance")

figure_test_2
# tiff(paste("total_models/plots/draft_all_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(figure_test_2)
# 
# # Closing the graphical device
# dev.off()

```

##Derivatives calculation

We calculated derivatives for each species per lake.

```{r}

# Derivatives() will ignore any random effect smooths it encounters in object.

# Table of pecies that were recorded with electro, fishbase or eawag, following Table 16 in Projet Lac Synthesis Report. 
# Plus the minimum and maximum temp in each lake -> needed for the new data for 
# calculating derivatives

species_lake <- read_xlsx("species_occurrences_lakes.xlsx") 
str(species_lake)

species_lake$fLake <- as.factor(species_lake$Lake)
species_lake$fProtocol <- as.factor(species_lake$Protocol)
species_lake$Species <- as.factor(species_lake$Species)

#Loading all four dfs to calculate derivatives

df_binomial_gam <- readRDS("data_frame_models/df_binomial_gam")
df_binomial_gam$fLake <- as.factor(df_binomial_gam$Lake)
df_binomial_gam$fProtocol <- as.factor(df_binomial_gam$Protocol)

df_abundance_gam <- readRDS("data_frame_models/df_abundance_gam")
df_abundance_gam$fProtocol <- as.factor(df_abundance_gam$Protocol)
df_abundance_gam$fLake <- as.factor(df_abundance_gam$Lake)

df_binomial_re <- readRDS("data_frame_models/df_binomial_re")
df_binomial_re$fLake <- as.factor(df_binomial_re$Lake)
df_binomial_re$fProtocol <- as.factor(df_binomial_re$Protocol)

df_abundance_re <- readRDS("data_frame_models/df_abundance_re")
df_abundance_re$fLake <- as.factor(df_abundance_re$Lake)
df_abundance_re$fProtocol <- as.factor(df_abundance_re$Protocol)

```

###Temp models
```{r}

# Now four different loops are run to calculate derivatives of each species 
# in each lake

# Model 1: Binomial data without random intercepts

species_list <- df_binomial_gam |>
  distinct(Species) |>
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

# loop through the list of species
for (i in species_list) {
  data <- df_binomial_gam |>
    filter(Species == i)

  lake_data <- species_lake |>
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'), family = binomial)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  # loop through all lakes where the species is present, lake list is based on the information from
  # Projet Lac
  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    # using minimum and maximum temp in the lake for derivatives
     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
                                             to = max(data_lake$temp, na.rm = TRUE), length = 200),
                       fProtocol = factor("VERT"))

     # calculate derivatives
     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
       mutate(fLake = factor(j)) |>
       mutate(species = factor(i))
       # rename(temp = mean_last_7days)

     saveRDS(derivatives, paste0("gams_testing_1/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

# Model 2.1: Abundance data and no random intercept for lake -> but binomial
# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
# run with binomial.

species_list <- df_abundance_gam |> 
  # binomial one
  filter(Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_gam |> 
    filter(Species == i)

  # binomial family instead of ZIP
  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
                           s(fProtocol, bs = 're'), family = binomial())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i))

    saveRDS(derivatives, paste0("gams_testing_1/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 2.2: Abundance data and no random intercept for lake -> ZIP
# all other species in this group run with ZIP

species_list <- df_abundance_gam |>
  filter(!Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_gam |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
                         family = ziP())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i))

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

```

```{r}

# Model 3: Binomial data and random intercept for lake

species_list <- df_binomial_re |> 
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_binomial_re |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
                         + s(fProtocol, bs = 're'), family = binomial)

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |> 
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    # take all lakes where the species is present and take random one for 
    # the data included in derivatives()
    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"), fLake = factor(random_lake))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i))

    saveRDS(derivatives, paste0("gams_testing_1/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

```

```{r}

# Model 4.1: Abundance data and random intercept for lake, complex models

species_list <- df_abundance_re |> 
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_re |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
                         +  s(fProtocol, bs = 're'), family = ziP())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |> 
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fLake = factor(random_lake), fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i))

    saveRDS(derivatives, paste0("gams_testing_1/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

```

```{r}

# save all the derivatives in one file

df_deriv <- list.files(path = "gams_testing_1/derivatives", pattern = ".rds", full.names = TRUE) |> 
  map_dfr(readRDS)

saveRDS(df_deriv, "gams_testing_1/df_deriv_all.rds")

```

##Calculating thermal response diversity

```{r}
# Function provided by Ross et al. "How to measure response diversity" used to calculate
# response diversity

# read in function
source(here("functions.R"))

# load derivatives data
all_models_derivatives <- readRDS("gams_testing_1/df_deriv_all.rds")

all_deriv <- as_tibble(all_models_derivatives)

str(all_deriv)

levels(all_deriv$species)

#Calculating response diversity for each lake community

all_deriv$fLake <- as.character(all_deriv $fLake)

lakes_list <- all_deriv |> 
  distinct(fLake) |> 
  pull(fLake)

str(lakes_list)

lakes_list <- sort(lakes_list)

species_overview <- tibble()

all_deriv$species <- as.character(all_deriv$species)
all_deriv$species <- as.factor(all_deriv$species)

str(all_deriv )

# Loop to get response diversity measures for each lake 
# loop through list of lakes and calculate response diversity in the lake communities
for (i in lakes_list){

  data <- all_deriv |>
    select(mean_last_7days, fLake, .derivative, species) |> 
    filter(fLake == i)

  df_resp_div <- data |>
    pivot_wider(
      names_from = species,
      values_from = .derivative)

  # apply the function from Ross et al. 
  # Dissimilarity
  df_resp_div$rdiv <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = F)
  # Divergence
  df_resp_div$sign <- apply(df_resp_div[,-(1:2), drop = FALSE], 1, resp_div, sign_sens = T)
  # Median Dissimilarity
  df_resp_div$Med <- median(df_resp_div$rdiv)

  # save response diversity metrics per lake
  saveRDS(df_resp_div, paste0("gams_testing_1/resp_div/df_resp_div_", i, ".rds"))

}

# save the response diversity per lake into one file
resp_div_no_excl <- list.files(path = "gams_testing_1/resp_div", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS) |>
  relocate(rdiv, Med, sign, .after = mean_last_7days)

saveRDS(resp_div_no_excl,"gams_testing_1/resp_div_all.rds")
```

#Drivers of thermal response diversity

Preparing a df with the information on past eutrophication, species richness, and species categorization in each lake.
```{r}
# Df with properties of lakes
lake_info <- read_xlsx("lake_info.xlsx")

# select the columns we use
lake_selection <- lake_info |> 
  select(Lake, Phos_max, Max_depth, Lake_area, Species_loss, Phos_max_Vonlanthen)

# information about categorization in each lake for the subgroups
species_endemism <- read_excel("species_category_per_lake.xlsx") |> 
  rename(endemism = detail_category)

# re-calculate species richness per lake
species_endemism_short <- species_endemism |> 
  # filter(species != "Lepomis_gibbosus") |> 
  select(-sum_species) |> 
  group_by(fLake) |> 
  mutate(sum_species = sum(num_species))

species_endemism_short$endemism <- as.factor(species_endemism_short$endemism)

str(species_endemism_short)

# adding column with number of species per category in each lake
df_species_endemism <- species_endemism_short |> 
  select(-species, - num_species) |> 
  add_count(fLake, endemism) |> 
  rename(count = n) |> 
  distinct(fLake, count, sum_species, endemism)

str(df_species_endemism)

df_species_endemism$count <- as.numeric(df_species_endemism$count)

# making it wide 
df_species_endemism_long <- df_species_endemism |> 
  pivot_wider(names_from = endemism, values_from = count) |> 
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) |> 
  rename(Lake = fLake)

str(df_species_endemism_long)

# merge endemism info and lake info

endemism_lake <- merge(df_species_endemism_long, lake_selection)

# Read in data on response diversity metrics in each lake
resp_div_all <- readRDS("gams_testing_1/resp_div_all.rds") |> 
  select(mean_last_7days:fLake)

# select response diversity columns we need
rdiv <- resp_div_all |>  
  select(mean_last_7days, fLake, rdiv, sign) |>
  drop_na()

# calculate mean and maximum values and rename Lake
df_rdiv <- rdiv |> 
  group_by(fLake) |> 
  mutate(mean_rdiv = mean(rdiv)) |> 
  mutate(max_rdiv = max(rdiv)) |> 
  mutate(mean_sign = mean(sign)) |> 
  mutate(max_sign = max(sign)) |> 
  distinct(fLake, mean_rdiv, max_rdiv, mean_sign, max_sign) |> 
  rename(Lake = fLake)

# combine lake properties, species subgroups and response diversity

df_lm <- merge(df_rdiv, endemism_lake)

str(df_lm)

# phosphorus values as numeric
df_lm$Phos_max <- as.numeric(df_lm$Phos_max)
df_lm$Phos_max_Vonlanthen <- as.numeric(df_lm$Phos_max_Vonlanthen)

```

##Fig. 3a
Dissimilarity, species richness and categories

```{r}
# Multiple linear regression for mean dissimilarity explained by species richness and past eutrophication: 
# lm(mean_rdiv  ~ species richness + eutrophication)
# Predict manually with new data eutrophication value fixed at mean and species richness min() to max()
# do the same the other way around: mean_rdiv  ~  eutrophication + species richness, with species richness fixed at mean and eutrophication min() to max()species richness +
#   prepare these two plots

# Phosphorus data on Poschiavo is not available
df_lm_excl <- df_lm |> 
  filter(Lake != "Poschiavo")

# first lm for species richness ~ maximum phosphorus

lm1 <- lm(mean_rdiv ~ sum_species + Phos_max, data = df_lm_excl)

# checking model assumptions
summary(lm1)
shapiro.test(resid(lm1))
lmtest::bptest(lm1)
plot(lm1)

# model assumptions validated

# manually predict data
new_data <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))

# predict lm
prediction_lm1 <- predict.lm(lm1, newdata = new_data, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model1 <- cbind(new_data, prediction_lm1) |> 
  mutate(category ="color")

mycolor = "black"

# pltting the first linear regression
lm1_plot <- bind_model1 |> 
  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean dissimilarity") +
  xlab("Species richness") +
  # ylim(1,3) +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

lm1_plot

```

##Figure 3b
```{r}
# Lms for all categories: endemic, translcated, non_native and non_endemic_native
# with data on all lakes
# Endemic
lm_endemic <- lm(mean_rdiv ~ endemic, data = df_lm)

# checking model assumptions
summary(lm_endemic)
shapiro.test(resid(lm_endemic))
lmtest::bptest(lm_endemic)
plot(lm_endemic)

new_data_a <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),length = 15))

prediction_endemic <- predict.lm(lm_endemic, newdata = new_data_a, se.fit = TRUE, type = "response")

df_lm_endemic <- cbind(new_data_a, prediction_endemic) |> 
  mutate(category = factor("endemic")) |> 
  rename(species_richness = endemic)

# Non-native

lm_nn <- lm(mean_rdiv ~ non_native, data = df_lm)

# checking model assumptions
summary(lm_nn)
shapiro.test(resid(lm_nn))
lmtest::bptest(lm_nn)
plot(lm_nn)

new_data_b <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
                                   length = 15))

prediction_nn <- predict.lm(lm_nn, newdata = new_data_b, se.fit = TRUE, type = "response")

df_lm_nn <- cbind(new_data_b, prediction_nn)  |> 
  mutate(category = factor("non_native")) |> 
  rename(species_richness = non_native)

# Non-endemic native

lm_nne <- lm(mean_rdiv ~ native, data = df_lm)

# checking model assumptions
summary(lm_nne)
shapiro.test(resid(lm_nne))
lmtest::bptest(lm_nne)
plot(lm_nne)

new_data_c <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
                                      length = 15))

prediction_nne <- predict.lm(lm_nne, newdata = new_data_c, se.fit = TRUE, type = "response")

df_lm_nne <- cbind(new_data_c, prediction_nne)  |> 
  mutate(category = factor("non_endemic_native")) |> 
  rename(species_richness = native)

# Translocated (non-native to region) species

lm_trans <- lm(mean_rdiv ~ non_native_region, data = df_lm)

# checking model assumptions
summary(lm_trans)
shapiro.test(resid(lm_trans))
lmtest::bptest(lm_trans)
plot(lm_trans)

new_data_d <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
                                              length = 15))

prediction_trans <- predict.lm(lm_trans, newdata = new_data_d, se.fit = TRUE, type = "response")

df_lm_trans <- cbind(new_data_d, prediction_trans)  |> 
  mutate(category = factor("translocated")) |> 
  rename(species_richness = non_native_region)

# Combine all four lm predictions
df_all_lms <- rbind(df_lm_endemic, df_lm_nn, df_lm_nne, df_lm_trans)

mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
               "translocated" = "#F2AD63")

plot_all_lms <- df_all_lms|> 
  ggplot(aes(species_richness, fit, color = category, fill = category, linetype = category)) +  
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
  theme_classic(base_size = 16) +
  ylab("Mean dissimilarity") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolors,
                    aesthetics = c("color", "fill"),
                    labels = c("endemic", "non-native", "non-endemic native", "translocated")) +
   scale_linetype_manual(values = c("endemic" = "solid",
                                   "non_native" = "dashed",
                                   "non_endemic_native" = "solid",
                                   "translocated" = "dashed"), labels = c("endemic", "non-native", "non-endemic native", "translocated"))

plot_all_lms

# save as TIFF
# tiff(paste("total_models/plots/lm_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
# # plot(ggarrange(depth1, depth2, ncol = 2))
# # plot science discussion
# plot(plot_all_lms)
# 
# dev.off()

```

## Figure 3 Divergence

```{r}

# first lm for species richness ~ maximum phosphorus

lm3 <- lm(mean_sign ~ sum_species + Phos_max, data = df_lm_excl)

# checking model assumptions
summary(lm3)
shapiro.test(resid(lm3))
lmtest::bptest(lm3)
plot(lm3)

# model assumptions validated

# manually predict data
new_data3 <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))

# predict lm
prediction_lm3 <- predict.lm(lm3, newdata = new_data3, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model3 <- cbind(new_data3, prediction_lm3) |> 
  mutate(category ="color")

lm3_plot <- bind_model3 |> 
  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean divergence") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

# Lms for all categories: endemic, translcated, non_native and non_endemic_native
# with data on all lakes
# Endemic
deriv_lm1 <- lm(mean_sign ~ endemic, data = df_lm)

# checking model assumptions
summary(deriv_lm1)
shapiro.test(resid(deriv_lm1))
lmtest::bptest(deriv_lm1)
plot(deriv_lm1)

new_data_1 <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
                                     length = 15))

prediction_endemic_1 <- predict.lm(deriv_lm1, newdata = new_data_1, se.fit = TRUE, type = "response")

df_deriv_endemic <- cbind(new_data_1, prediction_endemic_1) |> 
  mutate(category = factor("endemic")) |> 
  rename(species_richness = endemic)

# Non-native

deriv_lm2 <- lm(mean_sign ~ non_native, data = df_lm)

# checking model assumptions
summary(deriv_lm2)
shapiro.test(resid(deriv_lm2))
lmtest::bptest(deriv_lm2)
plot(deriv_lm2)

new_data_2 <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
                                   length = 15))

prediction_2 <- predict.lm(deriv_lm2, newdata = new_data_2, se.fit = TRUE, type = "response")

df_deriv_nn <- cbind(new_data_2, prediction_2)  |> 
  mutate(category = factor("non_native")) |> 
  rename(species_richness = non_native)

# Non-endemic native

deriv_lm3 <- lm(mean_sign ~ native, data = df_lm)

# checking model assumptions
summary(deriv_lm3)
shapiro.test(resid(deriv_lm3))
lmtest::bptest(deriv_lm3)
plot(deriv_lm3)

new_data_3 <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
                                      length = 15))

prediction_3 <- predict.lm(deriv_lm3, newdata = new_data_3, se.fit = TRUE, type = "response")

df_deriv_nne <- cbind(new_data_3, prediction_3)  |> 
  mutate(category = factor("non_endemic_native")) |> 
  rename(species_richness = native)

# Translocated (non-native to region) species

deriv_lm4 <- lm(mean_sign ~ non_native_region, data = df_lm)

# checking model assumptions
summary(deriv_lm4)
shapiro.test(resid(deriv_lm4))
lmtest::bptest(deriv_lm4)
plot(deriv_lm4)

new_data_4 <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
                                              length = 15))

prediction_4 <- predict.lm(deriv_lm4, newdata = new_data_4, se.fit = TRUE, type = "response")

df_deriv_trans <- cbind(new_data_4, prediction_4)  |> 
  mutate(category = factor("translocated")) |> 
  rename(species_richness = non_native_region)

# Combine all four lm predictions
df_deriv_lms <- rbind(df_deriv_endemic, df_deriv_nn, df_deriv_nne, df_deriv_trans)

mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
               "translocated" = "#F2AD63")

lm_divergence_all <- df_deriv_lms|> 
  ggplot(aes(species_richness, fit, color = category, fill = category)) +
  geom_line(linewidth = 1, linetype = "dashed") +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
  theme_classic(base_size = 16) +
  ylab("Mean divergence") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolors,
                    aesthetics = c("color", "fill"), 
                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
                    name = NULL)

lm_divergence_all

# save as TIFF
# tiff(paste("total_models/plots/lm_divergence_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
# # plot(ggarrange(depth1, depth2, ncol = 2))
# # plot science discussion
# plot(lm_divergence_all)
# 
# dev.off()
```
## Figure 4a Log
```{r}

# LOG
df_lm_4 <- df_lm |> 
  filter(!Lake %in% c("Poschiavo", "Joux", "Maggiore", "Lugano"))

# add log 
df_lm_4 <- df_lm_4 |> 
  mutate(P_max_log = log(Phos_max_Vonlanthen))

lm_species_loss_log <- lm(Species_loss ~ P_max_log, data = df_lm_4)

# checking model assumptions
summary(lm_species_loss_log)
shapiro.test(resid(lm_species_loss_log))
lmtest::bptest(lm_species_loss_log)
plot(lm_species_loss_log)

# new data
new_data_species_loss_log <- tibble(P_max_log = seq(from = min(df_lm_4$P_max_log), to = max(df_lm_4$P_max_log), length = 15))

# predict model
prediction_species_loss_log <- predict.lm(lm_species_loss_log, newdata = new_data_species_loss_log, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model_species_loss_log <- cbind(new_data_species_loss_log, prediction_species_loss_log) |> 
  mutate(category ="color")

# Plot Fig. 3
"#78BE63"
mycolor <- c("color" = "black")
# mycolor <- c("color" = "#A0E66A")
"98D47C"

test <- bind_model_species_loss_log|> 
  ggplot(aes(P_max_log, fit, color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Species loss") +
  xlab("Maximum total phosphorus (µg/L)") +
  # ylim(1,3) +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

test
```
## Figure 4a
```{r}

df_lm_4 <- df_lm |> 
  filter(!Lake %in% c("Poschiavo", "Joux", "Maggiore", "Lugano"))

# add log 
df_lm_4 <- df_lm_4 |> 
  mutate(P_max_log = log(Phos_max_Vonlanthen))

lm_species_loss <- lm(Species_loss ~ Phos_max_Vonlanthen, data = df_lm_4)

# checking model assumptions
summary(lm_species_loss)
shapiro.test(resid(lm_species_loss))
lmtest::bptest(lm_species_loss)
plot(lm_species_loss)

# new data
new_data_species_loss <- tibble(Phos_max_Vonlanthen = seq(from = min(df_lm_4$Phos_max_Vonlanthen), to = max(df_lm_4$Phos_max_Vonlanthen), length = 15))

# predict model
prediction_species_loss <- predict.lm(lm_species_loss, newdata = new_data_species_loss, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model_species_loss <- cbind(new_data_species_loss, prediction_species_loss) |> 
  mutate(category ="color")

# Plot Fig. 3
"#78BE63"
mycolor <- c("color" = "black")
# mycolor <- c("color" = "#A0E66A")
"98D47C"

bind_model_species_loss|> 
  ggplot(aes(Phos_max_Vonlanthen, fit, color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  geom_point(data = df_lm_4, aes(x = Phos_max_Vonlanthen, y = Species_loss), inherit.aes = FALSE, color = "black", size = 3) +
  theme_classic(base_size = 20) +
  ylab("Species loss") +
  xlab("Maximum total phosphorus (µg/L)") +
  # ylim(1,3) +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

```

## Figure 4b
```{r}
# second lm for maximum phosphorus ~ species richness
lm2 <- lm(mean_rdiv ~ Phos_max + sum_species, data = df_lm_excl)

# checking model assumptions
summary(lm2)
shapiro.test(resid(lm2))
lmtest::bptest(lm2)
plot(lm2)

# new data
new_data2 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))

# predict model
prediction_lm2 <- predict.lm(lm2, newdata = new_data2, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model2 <- cbind(new_data2, prediction_lm2) |> 
  mutate(category ="color")

# Plot Fig. 3
"#78BE63"
mycolor <- c("color" = "black")
# mycolor <- c("color" = "#A0E66A")
"98D47C"

lm2_plot <- bind_model2 |> 
  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean dissimilarity") +
  xlab("Maximum total phosphorus (µg/L)") +
  # ylim(1,3) +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

lm2_plot
# save as TIFF
# tiff(paste("total_models/plots/lm_species_richness_eutroph.tiff", sep = ""), units="in", width=12, height=5, res=300)
# 
# plot(figure_3)
# 
# # Closing the graphical device
# dev.off()

```

##Figure 4c
```{r}

lm_test <- lm(mean_rdiv ~ Species_loss, data = df_lm_4)

# checking model assumptions
summary(lm_test)
shapiro.test(resid(lm_test))
lmtest::bptest(lm_test)
plot(lm_test)

# new data
new_data_test <- tibble(Species_loss = seq(from = min(df_lm_4$Species_loss), to = max(df_lm_4$Species_loss), length = 15))

# predict model
prediction_test <- predict.lm(lm_test, newdata = new_data_test, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model_test <- cbind(new_data_test, prediction_test) |> 
  mutate(category ="color")

# Plot Fig. 3
"#78BE63"
mycolor <- c("color" = "black")
# mycolor <- c("color" = "#A0E66A")
"98D47C"
